---
layout:  page
title: "TCGA Data Download"
comments:  true
published:  true
author: "Marcin Kosiński"
date: 2016-02-24 00:45:00
categories: [RTCGA]
output:
  html_document:
    mathjax:  default
    fig_caption:  true
---

{:toc}

```{r include=FALSE}
library(knitr)
opts_chunk$set(comment="", message=FALSE, warning = FALSE, 
               tidy.opts=list(keep.blank.line=TRUE, width.cutoff=150),
							 options(width=150), eval = TRUE)
```

```{r}
library(RTCGA)
```

# Cohorts Names and Number of Cases

The Cancer Genome Atlas provides data via [`Broad GDAC Firehose`](http://gdac.broadinstitute.org/). The number of cases in the most popular datasets can be checked with the following code that is based on the [`Broad GDAC Firehose`](http://gdac.broadinstitute.org/).

```{r, results='asis'}
cat(knitr::kable(infoTCGA()), sep = "\n")
```


Furthermore `infoTCGA()` enables to extract possible cohorts names from TCGA Study. **Cohorts' names stand for abbreviations of real names of cancer types**

```{r}
library(magrittr)
(cohorts <- infoTCGA() %>% 
rownames() %>% 
   sub('-counts', '', x=.))
```


