---
layout:  page
title: "Visualizations"
comments:  true
published:  true
author: "Marcin Kosiński"
date: 2016-02-24 00:45:00
categories: [RTCGA]
output:
  html_document:
    mathjax:  default
    fig_caption:  true
    toc: true
    fig_width: 12
---

* `pcaTCGA()`
* `boxplotTCGA()`
* `kmTCGA()`
* `heatmapTCGA()`
{:toc}


```{r include=FALSE}
library(knitr)
opts_chunk$set(
	comment = "",
	fig.width = 12, 
	message = FALSE,
	warning = FALSE,
	tidy.opts = list(
		keep.blank.line = TRUE,
		width.cutoff = 150
		),
	options(width = 150),
	eval = TRUE
)
```

```{r}
library(dplyr)
library(RTCGA)
library(ggplot2)
```

# `pcaTCGA()`

## Biplot of RNASeq - Plot of 2 main componens of Principal Components Analysis

```{r}
## RNASeq expressions
library(RTCGA.rnaseq)
# library(dplyr) if did not load at start
expressionsTCGA(BRCA.rnaseq, OV.rnaseq, HNSC.rnaseq) %>%
 	dplyr::rename(cohort = dataset) %>%	
 	filter(substr(bcr_patient_barcode, 14, 15) == "01") -> BRCA.OV.HNSC.rnaseq.cancer
pcaTCGA(BRCA.OV.HNSC.rnaseq.cancer, "cohort") -> pca_plot
plot(pca_plot)
```

# `boxplotTCGA()`

## Boxplots of logarithm of MET gene RNASeq expression

```{r}
library(RTCGA.rnaseq)
# perfrom plot
# library(dplyr) if did not load at start
expressionsTCGA(
	ACC.rnaseq,
	BLCA.rnaseq,
	BRCA.rnaseq,
	OV.rnaseq,
	extract.cols = "MET|4233"
	) %>%
 	dplyr::rename(
 	cohort = dataset,
 	MET = `MET|4233`
 	) %>%  #cancer samples
 	filter(
 	substr(bcr_patient_barcode, 14, 15) == "01" 
 	) -> ACC_BLCA_BRCA_OV.rnaseq
boxplotTCGA(
	ACC_BLCA_BRCA_OV.rnaseq,
	"reorder(cohort,log1p(MET), median)",
	"log1p(MET)",
	xlab = "Cohort Type",
	ylab = "Logarithm of MET",
	legend.title = "Cohorts",
	legend = "bottom"
	) -> boxplot1
plot(boxplot1)
```

### Facet example

```{r}
library(RTCGA.mutations)
# library(dplyr) if did not load at start
mutationsTCGA(
	BRCA.mutations,
	OV.mutations,
	ACC.mutations,
	BLCA.mutations
	) %>% 
	filter(Hugo_Symbol == 'TP53') %>%
	filter(substr(bcr_patient_barcode, 14, 15) == 
	"01") %>% # cancer tissue
 	mutate(bcr_patient_barcode = 
 	substr(bcr_patient_barcode, 1, 12)) ->
	ACC_BLCA_BRCA_OV.mutations
 
mutationsTCGA(
	BRCA.mutations,
	OV.mutations,
	ACC.mutations,
	BLCA.mutations
) -> ACC_BLCA_BRCA_OV.mutations_all
 
 ACC_BLCA_BRCA_OV.rnaseq %>%
 	mutate(bcr_patient_barcode = 
 	substr(bcr_patient_barcode, 1, 15)) %>%
 	filter(
 	bcr_patient_barcode %in% 
 	substr(ACC_BLCA_BRCA_OV.mutations_all$bcr_patient_barcode, 1, 15)
 	) %>%
 # took patients for which we had any mutation information
 # so avoided patients without any information about mutations
 	mutate(bcr_patient_barcode = 
 	substr(bcr_patient_barcode, 1, 12)) %>%
 # strin_length(ACC_BLCA_BRCA_OV.mutations$bcr_patient_barcode) == 12
 	left_join(
 		ACC_BLCA_BRCA_OV.mutations,
 		by = "bcr_patient_barcode"
 		) %>% #joined only with tumor patients
 	mutate(TP53 = 
 	ifelse(!is.na(Variant_Classification), "Mut", "WILD")
 	) %>%
 	select(cohort, MET, TP53) -> 
 	ACC_BLCA_BRCA_OV.rnaseq_TP53mutations
 
boxplotTCGA(
	ACC_BLCA_BRCA_OV.rnaseq_TP53mutations,
	"reorder(cohort,log1p(MET), median)",
	"log1p(MET)",
	xlab = "Cohort Type",
	ylab = "Logarithm of MET",
	legend.title = "Cohorts",
	legend = "bottom",
	facet.names = c("TP53")
	) -> boxplo2
plot(boxplo2)
boxplotTCGA(
	ACC_BLCA_BRCA_OV.rnaseq_TP53mutations,
	"reorder(cohort,log1p(MET), median)",
	"log1p(MET)",
	xlab = "Cohort Type",
	ylab = "Logarithm of MET",
	legend.title = "Cohorts",
	legend = "bottom",
	fill = c("TP53")
	) -> boxplot3
plot(boxplot3)
boxplotTCGA(
	ACC_BLCA_BRCA_OV.rnaseq_TP53mutations,
	"reorder(TP53,log1p(MET), median)",
	"log1p(MET)",
	xlab = "Cohort Type",
	ylab = "Logarithm of MET",
	legend.title = "Cohorts",
	legend = "bottom",
	fill = c("cohort")
	) -> boxplot4
plot(boxplot4)
```

# `kmTCGA()`

## Kaplan-Meier estimates of survival curves for BRCA and OV cancer types and mutations in gene TP53

```{r}
library(RTCGA.mutations)
# library(dplyr) if did not load at start
library(survminer)
mutationsTCGA(BRCA.mutations, OV.mutations) %>%
 	filter(Hugo_Symbol == 'TP53') %>%
 	filter(substr(bcr_patient_barcode, 14, 15) ==
 	"01") %>% # cancer tissue
 	mutate(bcr_patient_barcode =
 	substr(bcr_patient_barcode, 1, 12)) ->
	BRCA_OV.mutations
 
library(RTCGA.clinical)
survivalTCGA(
	BRCA.clinical,
	OV.clinical,
	extract.cols = "admin.disease_code"
	) %>%
 	dplyr::rename(disease = admin.disease_code) ->
	BRCA_OV.clinical

BRCA_OV.clinical %>%
 	left_join(
 		BRCA_OV.mutations,
 		by = "bcr_patient_barcode"
 		) %>%
 	mutate(TP53 =
 	ifelse(!is.na(Variant_Classification), "Mut","WILDorNOINFO")) ->
	BRCA_OV.clinical_mutations
 
BRCA_OV.clinical_mutations %>%
select(times, patient.vital_status, disease, TP53) -> BRCA_OV.2plot

kmTCGA(
	BRCA_OV.2plot,
	explanatory.names = c("TP53", "disease"),
	break.time.by = 400,
	xlim = c(0,2000)
	) -> km_plot
print(km_plot)
```

## Kaplan-Meier estimates of survival curves for BRAF gene

```{r}
archivist::aread('MarcinKosinski/coxphSGD/1a06') %>% 
	do.call(rbind, . ) %>%
	kmTCGA(
		explanatory.names = "BRAF",
		break.time.by = 1000,
		xlim = c(0, 5000)
		) -> km_plot2
print(km_plot2)
```

# `heatmapTCGA()`

## Heatmap of medians of ZNF500 gene divided on cohort type and MET gene quantiles


```{r}
library(RTCGA.rnaseq)
# perfrom plot
# library(dplyr) if did not load at start
expressionsTCGA(
	ACC.rnaseq,
	BLCA.rnaseq,
	BRCA.rnaseq,
	OV.rnaseq,
	extract.cols = 
		c("MET|4233",
		"ZNF500|26048",
		"ZNF501|115560")
	) %>%
	dplyr::rename(cohort = dataset,
	MET = `MET|4233`) %>%
	#cancer samples
	filter(substr(bcr_patient_barcode, 14, 15) ==
	"01") %>%
	mutate(MET = cut(MET,
	 round(quantile(MET, probs = seq(0,1,0.25)), -2),
	 include.lowest = TRUE,
	 dig.lab = 5)) -> ACC_BLCA_BRCA_OV.rnaseq

ACC_BLCA_BRCA_OV.rnaseq %>%
	select(-bcr_patient_barcode) %>%
	group_by(cohort, MET) %>%
	summarise_each(funs(median)) %>%
	mutate(ZNF500 = round(`ZNF500|26048`),
	ZNF501 = round(`ZNF501|115560`)) ->
	ACC_BLCA_BRCA_OV.rnaseq.medians
heatmapTCGA(ACC_BLCA_BRCA_OV.rnaseq.medians,
	"cohort", "MET", "ZNF500",
	title = "Heatmap of ZNF500 expression")
```

### Facet examples with mutations datasets

```{r}
## facet example
library(RTCGA.mutations)
# library(dplyr) if did not load at start
mutationsTCGA(
	BRCA.mutations,
	OV.mutations,
	ACC.mutations,
	BLCA.mutations
	) %>%
	filter(Hugo_Symbol == 'TP53') %>%
	filter(substr(bcr_patient_barcode, 14, 15) ==
	"01") %>% # cancer tissue
	mutate(bcr_patient_barcode =
	substr(bcr_patient_barcode, 1, 12)) ->
	ACC_BLCA_BRCA_OV.mutations

mutationsTCGA(
	BRCA.mutations,
	OV.mutations,
	ACC.mutations,
	BLCA.mutations
	) -> ACC_BLCA_BRCA_OV.mutations_all

ACC_BLCA_BRCA_OV.rnaseq %>%
	mutate(bcr_patient_barcode =
	substr(bcr_patient_barcode, 1, 15)) %>%
	filter(bcr_patient_barcode %in%
substr(ACC_BLCA_BRCA_OV.mutations_all$bcr_patient_barcode, 1, 15)) %>% 
# took patients for which we had any mutation information
# so avoided patients without any information about mutations
	mutate(bcr_patient_barcode =
	substr(bcr_patient_barcode, 1, 12)) %>%
# strin_length(ACC_BLCA_BRCA_OV.mutations$bcr_patient_barcode) == 12
	left_join(ACC_BLCA_BRCA_OV.mutations,
	by = "bcr_patient_barcode") %>% #joined only with tumor patients
	mutate(TP53 = 
	ifelse(!is.na(Variant_Classification), "Mut", "WILD")
	) %>%
	select(-bcr_patient_barcode, -Variant_Classification,
	-dataset, -Hugo_Symbol) %>% 
	group_by(cohort, MET, TP53) %>% 
	summarise_each(funs(median)) %>% 
	mutate(ZNF501 = round(`ZNF501|115560`)) ->
	ACC_BLCA_BRCA_OV.rnaseq_TP53mutations_ZNF501medians

heatmapTCGA(
	ACC_BLCA_BRCA_OV.rnaseq_TP53mutations_ZNF501medians,
	"cohort",
	"MET",
	fill = "ZNF501",
	facet.names = "TP53",
	title = "Heatmap of ZNF501 expression"
)

heatmapTCGA(
	ACC_BLCA_BRCA_OV.rnaseq_TP53mutations_ZNF501medians,
	"TP53",
	"MET",
	fill = "ZNF501",
	facet.names = "cohort",
	title = "Heatmap of ZNF501 expression"
)

heatmapTCGA(
	ACC_BLCA_BRCA_OV.rnaseq_TP53mutations_ZNF501medians,
	"TP53",
	"cohort",
	fill = "ZNF501",
	facet.names = "MET",
	title = "Heatmap of ZNF501 expression"
)
```

