---
title: "Using `RTCGA` package to download `r name` data that are included in RTCGA.`r name`.`r gsub('-', '', releaseDate)` package"
subtitle: "Date of datasets release: `r releaseDate`"
author: "`r author`"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using RTCGA to download `r name` data as included in RTCGA.`r name`.`r gsub('-', '', releaseDate)`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(
	comment = "",
	message = FALSE,
	warning = FALSE,
	tidy.opts = list(
		keep.blank.line = TRUE,
		width.cutoff= 150),
	options(width= 150),
	eval = TRUE
)
```

# RTCGA package

> The Cancer Genome Atlas (TCGA) Data Portal provides a platform for researchers to search, download, and analyze data sets generated by TCGA. It contains clinical information, genomic characterization data, and high level sequence analysis of the tumor genomes. The key is to understand genomics to improve cancer care.

`RTCGA` package offers download and integration of the variety and volume of TCGA data using patient barcode key, what enables easier data possession. This may have a benefcial infuence on  development of science and improvement of patients' treatment. `RTCGA` is an open-source R package, available to download from Bioconductor 

```{r}
# source("http://bioconductor.org/biocLite.R")
# biocLite("RTCGA")
```

or from github
```{r}
# if (!require(devtools)) {
#    install.packages("devtools")
#    require(devtools)
# }
# biocLite("RTCGA/RTCGA")
```

Furthermore, `RTCGA` package transforms TCGA data into form which is convenient to use in R statistical package. Those data transformations can be a part of statistical analysis pipeline which can be more reproducible with `RTCGA`.

Use cases and examples are shown in `RTCGA` packages vignettes:
```{r}
# browseVignettes("RTCGA")
```

# How to download `r name` data to gain the same datasets as in RTCGA.`r name`.`r gsub('-', '', releaseDate)` package?

There are many available date times of TCGA data releases. To see them all just type:
```{r}
library(RTCGA)
library(magrittr)
# checkTCGA('Dates')
```

Version `r gsub('-', '', releaseDate)` of RTCGA.`r name`.`r gsub('-', '', releaseDate)` package contains `r name` datasets which were released `r releaseDate`. They were downloaded in the following way (which is mainly copied from [http://rtcga.github.io/RTCGA/](http://rtcga.github.io/RTCGA/):

## Available cohorts

All cohort names can be checked using:
```{r}
(cohorts <- infoTCGA() %>% 
   rownames() %>% 
   sub("-counts", "", x=.))
```

For all cohorts the following code downloads the `r name` data.

## Downloading `r name` files
```{r}
dir.create( "data2" ) # name of a directory in which data will be stored
sapply( cohorts, function(element){
tryCatch({
downloadTCGA( cancerTypes = element, 
							dataSet = dataSet,
              destDir = "data2", 
              date = releaseDate )},
error = function(cond){
   cat("Error: Maybe there weren't", name, " data for ", element, " cancer.\n")
}
)
})
```

## Reading downloaded `r name` dataset

### Shortening paths and directories 

```{r}
list.files( "data2") %>% 
   file.path( "data2", .) %>%
   file.rename( to = substr(.,start=1,stop=50))
```


### Removing `NA` files from data2 folder

If there were no `r name` data for some cohorts we should remove corresponding `NA` files.

```{r}
list.files( "data2") %>%
   file.path( "data2", .) %>%
   sapply(function(x){
      if (x == "data2/NA")
         file.remove(x)      
   })
```

### Paths to `r name` data

Below is the code that automatically assigns paths to files for all `r name` files for all available cohorts types downloaded to `data2` folder.

```{r}
cohorts %>%
	sapply(function(z){
		list.files("data2") %>%
			file.path("data2", .) %>%
			grep(paste0("_",z,"\\."), x = ., value = TRUE) %>%
			file.path(., list.files(.)) %>%
			grep("dataSetFile", x = ., value = TRUE) %>%
			assign(value = .,
						 x = paste0(z, ".",name,".path"),
						 envir = .GlobalEnv)
	})
```

### Reading `r name` data using `readTCGA`

Because of the fact that `r name` data are transposed in downloaded files, there has been prepared special function `readTCGA` to read (with data.table::fread) and transpose data automatically. Code is below

```{r}
ls() %>%
   grep(paste0(name,"\\.path"), x = ., value = TRUE) %>% 
   sapply(function(element){
      tryCatch({
         readTCGA(get(element, envir = .GlobalEnv),
               dataType = "dataType") -> read_file
         	
		     ## remove non-ASCII strings:
		     for( i in 1:ncol(read_file)){
		       read_file[, i] <- iconv(read_file[, i],
		                                    "UTF-8", "ASCII", sub="")
		     } 
         	
         assign(value = read_file,
                x = sub("\\.path", "", x = element),
                envir = .GlobalEnv )
      }, error = function(cond){
        cat(element)
      })
     invisible(NULL)
    }    
)
```

# Saving `r name` data to RTCGA.`r name`.`r gsub('-', '', releaseDate)` package


```{r}
grep( name, ls(), value = TRUE) %>%
   grep("path", x=., value = TRUE, invert = TRUE) %>%
   paste0( collapse="," ) -> use_data_input
   # ...	Unquoted names of existing objects to save
eval(parse(text=
   paste0("devtools::use_data(",use_data_input,",pkg='../',compress=\"xz\")")
))
```
